Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-02-11 papers/histogram/sad.tex
edits to paper

We use the average entropy error vs moves as a metric to compare
simulation runtimes and overall convergence. The overall accuracy
is determined by examining the fractional error of a particular method to
a precise reference system. For each simulation, the reference system
is chosen to be the final output of the SAMC simulation with the best
choice of $t_0$.
\begin{align}
\epsilon_\text{avg} = 1 - \bigg\lvert\frac{S_\text{method}}{S_\text{ref}}\bigg\rvert
\end{align}
Simulations that take more iterations to reach the same entropy error as a
different method take longer to converge. This behavior is seen in
Fig.~\ref{fig:N50-ff0.3-avg-error}.
\jpsays{We use a SAMC $t_0 = 10^3$ with a fixed energy range $S_{\max} = 120$ and $S_{\min} = 248$ as the reference.
This is noted because this is not strictly SAMC method. We use this as a reference to increase convergence time of the
reference.}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
  a fixed energy range SAMC $t_0 = 10^3$ serving as the reference.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-slow.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
  a fixed energy range SAMC $t_0 = 10^3$ from Fig.~\ref{fig:N50-ff0.3-avg-error}
  serving as the reference.}\label{fig:N50-ff0.3-avg-error-slow}
\end{figure}

@@

Our smaller system consists of 50 square-well atoms with a well width
of $\lambda=1.3$, in a volume corresponding to a filling fraction of
$\eta=0.3$.  For this simulation, we chose a minimum reduced
temperature of $1/3$, which corresponds to an interesting energy range
from -248 to -120.  The entropy of this system is shown in
Fig.~\ref{fig:entropy-cartoon} above, which shows that over this
energy range the entropy differs by 198.

Figures~\ref{fig:N50-ff0.3-avg-error}
and~\ref{fig:N50-ff0.3-avg-error-slow} show the average error in the entropy vs
moves for a system consisting of 50 atoms that are placed in a periodic
simple cubic unit cell with lattice constant chosen to give a filling
fraction of $\eta = 0.3$.  These particles are randomly inserted into
the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05\sigma$ and $\delta_0 = 0.005\sigma$ respectively). For
the $\eta = 30\%$ and $N = 50$ system, we explore the choice of
$\delta_0$ as a parameter.

The interesting range of energies for this system is from
{\color{blue}
$S_{\max} = 120$ and $S_{\min} = 248$ for this system.  The number of
important energy states for this system is then $N_S = S_{\min} - S_{\max} = 145$.
The reduced temperature of $1/3$ is chosen in order to examine the convergence
properties of SAD in a reasonable simulation time.

%As a side note, WL-TMMC does not converge well for $\delta_0 = 0.05\sigma$ while
%it does perform well for $\delta_0 = 0.005\sigma$. The WL portion of the algorithm
%spends nearly 100x longer for the smaller translation scale ensuring that the algorithm
%converges quickly upon transitioning to TMMC.
}

Smaller translation
scales cause the given method to take additional time to explore all
energies. Ideal methods should scale linearly with the $\delta_0$, that
is, a decrease in the translation scale should result in a proportional
increase in convergence time. As can be seen from
Fig.~\ref{fig:N50-ff0.3-avg-error-slow}, SAMC methods that rapidly
converged for $\delta = 0.05\sigma$ do not proportionaly converge for a
translation scale of $\delta = 0.005\sigma$. We see that SAD handles
arbitrary translation scales and proportionaly converges based on the
translation scale that is chosen.

>>>>

We use the average entropy error vs moves as a metric to compare
simulation runtimes and overall convergence. The overall accuracy
is determined by examining the fractional error of a particular method to
a precise reference system. For each simulation, the reference system
is chosen to be the final output of the SAMC simulation with the best
choice of $t_0$.
\begin{align}
\epsilon_\text{avg} = 1 - \bigg\lvert\frac{S_\text{method}}{S_\text{ref}}\bigg\rvert
\end{align}
Simulations that take more iterations to reach the same entropy error as a
different method take longer to converge. This behavior is seen in
Fig.~\ref{fig:N50-ff0.3-avg-error}.
\jpsays{We use a SAMC $t_0 = 10^3$ with a fixed energy range $S_{\max} = 120$ and $S_{\min} = 248$ as the reference.
This is noted because this is not strictly SAMC method. We use this as a reference to increase convergence time of the
reference.}

\davidsays{Add this info to the section above.}  The SAMC simulations
computed the density of states for the entire range of possible
energies.  The SAD simulations determined the energy range of interest
dynamically as described above, based on a specified $T_{\min}$.  For
the WL and $1/t$-WL simulations, we constrained the simulation to
remain in the energy range corresponding to $T_{\min} < T < \infty$,
as determined by a previous SAMC simulation.  Thus the WL and $1/t$-WL
simulations were given extra information that in practice would not be
available without additional computational effort.

@@

\davidsays{The following paragraph duplicates some of what is written
  in the previous.  Please move things up as seems best!}

Our smaller system consists of 50 square-well atoms with a well width
of $\lambda=1.3$, in a volume corresponding to a filling fraction of
$\eta=0.3$.  For this simulation, we chose a minimum reduced
temperature of $1/3$, which corresponds to an interesting energy range
from -248 to -120.  The entropy of this system is shown in
Fig.~\ref{fig:entropy-cartoon} above, which shows that over this
energy range the entropy differs by 198, corresponding to a ratio of
$10^{86}$ between the highest and lowest density of states.

In order to explore the effect of simulation details on convergence,
we consider two values for the mean distance by which atoms are moved
during a Monte Carlo step.  We began with a reasonable displacement
distance of $\delta_0 = 0.05\sigma$, which corresponds to an
acceptance rate of 38\%.  We further ran simulations with a much
smaller displacement distance of $\delta_0 = 0.005\sigma$, which
resulted in an acceptance rate of 86\%, which we expected converge
more slowly.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-default.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
  a fixed energy range SAMC $t_0 = 10^3$ serving as the reference.}\label{fig:N50-ff0.3-avg-error}
\end{figure}
\begin{figure}
  \includegraphics[width=\columnwidth]{figs/s000/periodic-ww1_30-ff0_30-N50-entropy-error-slow.pdf}
  \caption{The average entropy error for each MC method is shown for $\eta = 0.3$ and $T_{\min} = 1/3$ with
  a fixed energy range SAMC $t_0 = 10^3$ from Fig.~\ref{fig:N50-ff0.3-avg-error}
  serving as the reference.}\label{fig:N50-ff0.3-avg-error-slow}
\end{figure}

Figure~\ref{fig:N50-ff0.3-avg-error} shows the average error in the
entropy as a function of time for this system with the reasonable
displacement distance of $\delta_0 = 0.05\sigma$.  The solid lines
represent the average of the absolute value of the error in the
entropy averaged over four simulations using different random number
seeds.  The range of average errors for each simulation is shown as a
shaded region around its mean error.  By the time $10^8$ moves have
been made all but the SAMC simulation with the shortest $t_0$ have
begun to converge as $1/\sqrt{t}$.  We then see the WL error begin to
saturate around $10^{10}$ moves.

Figure~\ref{fig:N50-ff0.3-avg-error-slow} shows the average error in
the entropy as a function of time for this system with the
unreasonably small displacement distance of $\delta_0 = 0.005\sigma$.
%
The smaller translation scale causes all methods to take additional
time to explore all energies. Based on random walk scaling, a ideal
method should scale roughly as $\delta_0^{-2}$ in the limit of small
$\delta_0$, that is, one order of magnitude in the displacement
distance should result in two order of magnitude increase in
convergence time. SAMC simulations with a $t_0$ value that rapidly
converged for $\delta = 0.05\sigma$ do not converge at all for a
translation scale of $\delta = 0.005\sigma$. SAD, WL, and $1/t$-WL
handle the shift in displacement distance and converge roughly as
expected.
